---
title: "BioCondition Field Operations Prioritisation Tool — Technical Documentation"
subtitle: "Priority Scoring Methodology & Implementation"
author: "Brodie Verrall, Gabrielle Lebbink"
date: "2026-07-17"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
    theme: flatly
    highlight: tango
  pdf_document:
    toc: true
    toc_depth: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

---

## Overview

This tool generates a prioritisation ranking for Regional Ecosystems (REs)
requiring BioCondition field assessments. REs are scored across several
dimensions — benchmark site deficit, vegetation extent decline, pre‑clearing
extent, remnant extent, SBC model deficits, benchmark reliability, and
benchmark requests — each weighted to reflect operational priority.

The output is a statewide and bioregional ranking, decile, and tier
classification (Very High, High, Medium, Low, Very Low). Spatial layers
(pre‑clearing and remnant GeoPackages) are annotated with the priority
attributes and exported alongside a CSV summary table.

---

## Pipeline Summary

| Section | Purpose |
|---------|---------|
| 0 | Environment setup: packages, directories |
| 1 | Import source data: RE extents, benchmarks, site records |
| 2 | Process RE extents: pre‑clearing & remnant area calculation |
| 3 | Build RE assessment dataset: join attributes |
| 4 | Benchmark information: selection, analogues, requests |
| 5 | Site counts: eligible QBEIS/QBERD sites, large‑tree data |
| **6** | **Generate Prioritisation Ranking ← core function** |
| 7 | Export CSV, GeoPackage outputs |

---

## Section 6 — Prioritisation Function

The function `calc_benchmark_priority()` performs all ranking logic.

### Function Signature

```r
calc_benchmark_priority(
  df, analogue_scores, sbc_defs, bm_reliability, re_codes,
  benchmark_target = 3,
  wt_deficit      = 0.175,   # benchmark site deficit
  wt_decline      = 0.125,   # proportional vegetation loss
  wt_extent       = 0.200,   # preclear extent
  wt_remnant      = 0.050,   # remnant extent
  wt_sbc          = 0.025,   # SBC modelling deficit
  wt_reliability  = 0.100,   # automated benchmark reliability
  wt_request      = 0.125,   # BM requests
  wt_return       = 0.250    # analogue influence (return score multiplier)
)
```

### Scoring Eligibility (`score_required`)

An RE receives a priority score only if at least one deficit remains:

```r
BM_effective = if_else(BM == "Y" & status == "draft", "N", BM)

score_required =
  (BM_effective == "Y" &  sbc_deficit > 0) |
  (BM_effective != "Y" & (benchmark_deficit > 0 | sbc_deficit > 0))
```

- **Benchmark available** → scored only if an SBC gap exists.
- **No benchmark (or draft status)** → scored if any site deficit or SBC gap
  exists. Benchmarks with `status = "draft"` are treated as not yet available,
  so the RE is still eligible for prioritisation.
- REs that already meet all criteria (effective benchmark present + no SBC
  gap) are excluded from ranking and receive `NA`.


### Factor Breakdown

Each factor is normalised using the internal `norm01()` function before
entering the weighted sum:

```r
norm01 <- function(x) {
  rng <- range(x, na.rm = TRUE)
  if ((rng[2] - rng[1]) == 0) return(rep(0, length(x)))
  (x - rng[1]) / (rng[2] - rng[1])
}
```

This produces a 0–1 scale. The reliability factor is then inverted
(`1 - scaled`) so that lower reliability yields a higher contribution.

#### 1. Benchmark Site Deficit — `wt_deficit = 0.175`

Target: 3 standard (or large‑tree) sites per RE.

```r
BM_effective = if_else(BM == "Y" & status == "draft", "N", BM)

benchmark_deficit = case_when(
  BM_effective == "Y"  ~ 0,
  TRUE                 ~ pmax(benchmark_target - benchmark_sites, 0)
)
```

Large‑tree data are required for wooded REs; non‑wooded REs use standard
site counts. If an effective benchmark already exists (published or
unpublished — but not draft), the deficit is set to 0. Draft benchmarks are
treated as not yet available.

#### 2. Vegetation Decline — `wt_decline = 0.125`

```r
decline_prop = if_else(Preclear_ha > 0, Difference_ha / Preclear_ha, 0)
```

The proportional loss of extent since pre‑clearing. Higher decline → greater
priority.

#### 3. Pre‑clearing Extent — `wt_extent = 0.200`

```r
extent_scaled = norm01(log1p(Preclear_ha))
```

Log‑transformed pre‑clearing area. Larger extents score higher, reflecting
the broader spatial impact.

#### 4. Remnant Extent — `wt_remnant = 0.050`

```r
remnant_scaled = norm01(log1p(Remnant_ha))
```

Log‑transformed remnant area. A small weight acknowledges that REs with very
little remaining area warrant attention.

#### 5. SBC Model Deficit — `wt_sbc = 0.025`

```r
  sbc_deficit = if_else(
    RE1 == "11.9.5" | benchmark_sites > 10, 0,
    pmax(coalesce(TD_ref_combined_deficit, 0), 0)
  )
```

The training‑data gap for State‑wide BioCondition modelling. Capped at 0
(minimum). Zeroed for RE `11.9.5` (east–west benchmark alignment artefact)
and for any RE with more than 10 benchmark‑quality sites
(`benchmark_sites > 10`), as these already have abundant field data. A small
weight reflects that SBC deficits are secondary to field benchmark needs.

#### 6. Benchmark Reliability — `wt_reliability = 0.100`

Missing reliability scores are handled in three tiers:

```r
has_reliability =
  !is.na(overall_reliability) | (RE1 %in% re_codes)

reliability_score = if_else(
  !is.na(overall_reliability), overall_reliability, 0
)

reliability_deficit_scaled = 1 - norm01(reliability_score)
```

| Case | Treatment |
|------|-----------|
| RE has a reliability score in `BM_automisation_data` | Used as-is |
| Valid RE (in `regional_ecosystem.csv`) missing from reliability data | `reliability_score = 0` → maximum priority contribution |
| Non‑standard RE code not in `regional_ecosystem.csv` | `has_reliability = FALSE` → term is zeroed out in `need_score` |

The reliability term is only included when `has_reliability = TRUE`.

#### 7. Benchmark Requests — `wt_request = 0.125`

```r
request_scaled = norm01(bm_requests_2526)
```

Count of pending benchmark requests for the 2025–26 cycle. A larger number
of requests increases priority.

#### 8. Analogue Influence (Return Score) — `wt_return = 0.250`

```r
analogue_influence = sum(analogue_weight)  # rank_1=8, rank_2=4, rank_3=2, rank_4=1
return_score = norm01(analogue_influence)
```

The "return" component — REs with strong analogue benchmarks provide
high information return for the effort. Applied as a multiplier rather than
an additive term.

---

### Final Priority Score Formula

```r
need_score =
  wt_deficit      * deficit_scaled +
  wt_decline      * decline_scaled +
  wt_extent       * extent_scaled +
  wt_remnant      * remnant_scaled +
  wt_sbc          * sbc_scaled +
  wt_reliability  * reliability_deficit_scaled +  # conditional on has_reliability
  wt_request      * request_scaled

priority_score = need_score * (1 + wt_return * return_score)
```

- **`need_score`** — additive weighted sum of all deficit/risk factors
- **`return_score`** — multiplicative boost for high‑analogue REs
- The final `priority_score` ranks REs statewide and within bioregions

---

### Weight Summary Table

| Factor | Weight | Direction | Rationale |
|--------|--------|-----------|-----------|
| Benchmark site deficit | 0.175 | ↑ deficit ↑ priority | Sites are the primary field need |
| Vegetation decline | 0.125 | ↑ decline ↑ priority | Threatened ecosystems |
| Pre‑clearing extent | 0.200 | ↑ extent ↑ priority | Landscape‑scale impact |
| Remnant extent | 0.050 | ↑ remnant ↑ priority | Small remnants need attention |
| SBC deficit | 0.025 | ↑ deficit ↑ priority | Modelling data gap |
| Benchmark reliability | 0.100 | ↓ reliability ↑ priority | Prioritise uncertain benchmarks |
| BM requests | 0.125 | ↑ requests ↑ priority | Operational demand signal |
| **Analogue return** | **0.250** | **(multiplier)** | Efficiency / information gain |

---

### RE Code Handling

To maximise matching between source datasets, RE codes are cleaned inside the
function before the reliability join:

```r
RE_lookup = case_when(
  str_detect(RE1, "/")                        ~ str_extract(RE1, "^[^/]+"),
  RE1 %in% c("11.9.5_eastern", "11.9.5_western") ~ "11.9.5",
  TRUE                                        ~ RE1
)
```

- **Compound REs** (e.g. `11.3.25/11.3.1`) — use the first code only
- **`11.9.5_eastern` / `11.9.5_west`** — map to base RE `11.9.5` for
  reliability lookup
- The SBC deficit for `11.9.5` is forced to 0 (east–west benchmark artefact)

---

## Outputs

| File | Contents |
|------|----------|
| `YYYYMMDD_BC_deficit.csv` | Tabular priority results with all attributes |
| `YYYYMMDD_RE_PC_v14_PRIORITISED.gpkg` | Pre‑clearing spatial layer with priority fields |
| `YYYYMMDD_RE_RE_v14_PRIORITISED.gpkg` | Remnant spatial layer with priority fields |

### Priority Fields

- `RE1`, `BR_code`, `LZ_code`, `VC_code` — ecosystem identifiers
- `Preclear_ha`, `Remnant_ha`, `Difference_ha` — extent metrics
- `benchmark_deficit`, `sbc_deficit`, `lt_deficit` — component deficits
- `priority_score`, `priority_rank`, `priority_decile`, `priority_tier` — statewide
- `priority_rank_BR`, `priority_decile_BR`, `priority_tier_BR` — bioregional
- `prioritisation_flag` — descriptive category with SBC gap annotation
- Biodiversity status, VMA class, protected area coverage, and more

---

## Dependencies

```{r deps, echo=FALSE}
packages <- c("dplyr", "tidyr", "stringr", "arrow", "sf", "cli", "lubridate")
cat(paste0("- ", packages, collapse = "\n"))
```

---

## Notes

- Run from the project root. The script automatically detects or creates a
  `project_data/` subdirectory.
- Verify input datasets (RE, BM, SBC) are current before each run.
- The `benchmark_target = 3` default reflects the standard BioCondition
  requirement of three representative sites per RE.
- Non‑standard RE codes (e.g. `af` suffixes, compound codes) that cannot be
  matched to the reliability table are excluded from the reliability
  component but still scored on remaining factors.

---

## Detailed Section Breakdown

### Section 0 — Environment Setup

The script activates the project working directory under `project_data/`
and ensures all required R packages (`dplyr`, `tidyr`, `stringr`, `arrow`,
`sf`, `cli`, `lubridate`) are installed and loaded. Input and output
subdirectories (`inputs/`, `outputs/`) are created if absent.

### Section 1 — Import Source Data

Seven tabular datasets and two spatial GeoPackages are loaded:

| File | Object | Contents |
|------|--------|----------|
| `All_summary_data.csv` | `all_data` | Site‑level QBEIS/QBERD data |
| `BENCHMARK_MASTER_LIST_v14.0.csv` | `bm_master` | Benchmark registry |
| `ranked_re_cleanlong_EDL(in).csv` | `bm_analogous` | Analogue RE rankings |
| `BM_automisation_data.csv` | `bm_reliability` | Automated reliability scores |
| `BM_requests_2526.csv` | `bm_requests` | Pending BM development requests |
| `20250904_SBC_trainingData_deficit.csv` | `sbc_defs` | SBC model training gaps |
| `regional_ecosystem.csv` | `re_v14` | RE metadata and attributes |
| `preclear_v14_3577.gpkg` | `re_pc_v14` | Pre‑clearing spatial polygons |
| `remnant_v14_3577.gpkg` | `re_re_v14` | Remnant spatial polygons |

Spatial geometries are dropped after reading to produce tabular extent
summaries for efficient joining.

### Section 2 — Process RE Extents

RE metadata is parsed by splitting the `NAME` field into bioregion (`BR_code`),
land zone (`LZ_code`), and vegetation community (`VC_code`) components.
Mapped extents from the REDD database are extracted from the
`MAPPED_EXTENTS` field.

Pre‑clearing and remnant spatial layers are grouped by `RE1` and aggregated
to produce:

- `Preclear_ha` — total pre‑clearing area (hectares)
- `Remnant_ha` — total remnant area (hectares)
- `Difference_ha` — Preclear_ha − Remnant_ha (clearing extent)
- `RE_parent` — base RE code with variant suffix stripped

### Section 3 — Build RE Assessment Dataset

The RE extent table is joined with ecosystem attributes from
`regional_ecosystem.csv`, including structural classification, short/long
descriptions, biodiversity status, VMA class, protected area coverage, and
habitat notes. This forms the base table `bc_re1` that accumulates all
subsequent joins.

### Section 4 — Benchmark Information

**4.1 Benchmark Selection** (`bm_selected`):
The Benchmark Master List is filtered to retain the most recent published
benchmark entry per RE. For REs with structural variance (multiple BM.RE
codes), all variant codes are concatenated. The `status` field (draft, final
unpublished, final published) is retained and used within the scoring
function to treat draft benchmarks as not yet available for prioritisation
purposes.

**4.2 Benchmark Requests** (`bm_requests_2526`):
Request counts from the 2025–26 cycle are joined to `bm_selected`. REs
without requests receive a count of `0L`.

**4.3 Benchmark Analogues** (`analogue_scores`):
The analogue dataset is processed by the helper function
`process_analogues()`, which assigns numeric weights to each analogue rank
(rank\_1 = 8, rank\_2 = 4, rank\_3 = 2, rank\_4 = 1) and sums them per RE
to produce `analogue_influence`.

**4.4 Benchmark Reliability**:
Handled within the scoring function in Section 6 (see Reliability breakdown
above).

### Section 5 — Generate Site Counts

Eligible sites are selected from `all_data` based on database origin and data
quality criteria:

- **QBEIS sites**: floristic sample grades A, B, or C and representative
  flag = 1.
- **QBERD sites**: site type matching "reference", "remnant", or "ref"
  (case‑insensitive), or representative flag = 1 with a CORVEG site type.

Two site‑count tables are produced:

| Table | Count Field | Description |
|-------|-------------|-------------|
| `site_counts_all` | `QBEIS_rep`, `QBERD_rep` | All eligible sites |
| `site_counts_large_trees` | `QBEIS_rep_LT`, `QBERD_rep_LT` | Sites with large‑tree data |

No‑match and unassignable RE codes are excluded. The two tables are
full‑joined and merged into `bc_re1` alongside the benchmark selection.

Rows with missing or empty `VC_code` are removed after these joins,
ensuring only valid REs proceed to scoring.

### Section 6 — Generate Prioritisation Ranking

This is the core of the tool. See the dedicated section above for the full
methodology covering:

- RE code cleaning (`RE_lookup`) for compound and variant codes
- `score_required` eligibility logic
- All eight factor calculations with their weights
- The two‑component `need_score` × `(1 + wt_return × return_score)` formula
- Statewide and bioregional ranking, decile, and tier assignment

### Section 7 — Generate Prioritisation Outputs

**7.1–7.2 — Tabular export**:
A predefined field list (`priority_fields`) is selected from the scored
table and written to `outputs/{date}_BC_deficit.csv`.

**7.3–7.4 — Spatial annotation**:
Pre‑clearing and remnant GeoPackages are read, geometries are cleaned (Z‑dimension
dropped, cast to `MULTIPOLYGON`), and priority attributes are joined by `RE1`.
Rows with missing `VC_code` are excluded.

**7.5–7.6 — Spatial export**:
Output fields are selected (spatial identifiers + all priority fields) and
written to:

- `outputs/{date}_RE_PC_v14_PRIORITISED.gpkg`
- `outputs/{date}_RE_RE_v14_PRIORITISED.gpkg`

Both GeoPackages are overwritten (`delete_dsn = TRUE`) to ensure clean
outputs on each run.
